---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# STExplorer

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/LefterisZ/STExplorer)](https://github.com/LefterisZ/STExplorer/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/LefterisZ/STExplorer)](https://github.com/LefterisZ/STExplorer/pulls)
<!-- badges: end -->

The goal of `STExplorer` is to utilise Geography's spatial data analysis techniques to perform spatially-aware analysis of spatial transcriptomics data. 

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `STExplorer` from [Bioconductor](http://bioconductor.org/) using the following code:

This option is still **NOT** available. Please use the GitHub installation from below. The package is Bioconductor-compatible and we are working towards submitting it to Bioconductor as soon as possible.

```{r 'install', eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE)) {
#     install.packages("BiocManager")
# }
# 
# BiocManager::install("STExplorer")
```

And the development version from [GitHub](https://github.com/LefterisZ/STExplorer) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("LefterisZ/STExplorer")
```
## Example

The package includes a comprehensive set of vignettes with a variety of examples and use cases. Please refer to the vignettes.

## Citation

Below is the citation output from using `citation('STExplorer')` in R. Please
run this yourself to check for any updates on how to cite __STExplorer__.

```{r 'citation', eval = requireNamespace('STExplorer')}
print(citation('STExplorer'), bibtex = TRUE)
```

Please note that the `STExplorer` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `STExplorer` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

